\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2020-08-25 sections/abstract.tex
Updates from Overleaf

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks that can help understand and control microbial community assembly in different biomes. 
        The pipeline is available as a Python package at \href{https://github.com/segrelab/MiCoNE}{https://github.com/segrelab/MiCoNE}
  }
\end{abstract}

>>>>

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of associations in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that can convert 16S sequencing data into a network of microbial associations.
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks that can help understand and control microbial community assembly in different biomes.
    The pipeline is available as a Python package at \href{https://github.com/segrelab/MiCoNE}{https://github.com/segrelab/MiCoNE}
  }
\end{abstract}

### 2020-08-25 sections/acknowledgments.tex
Updates from Overleaf

% TODO: Update this section
\section*{Acknowledgments}

We are grateful to members of the Segrè lab for helpful discussions and for feedback on the manuscript. This work was partially funded by grants from the National Institutes of Health (National Institute of General Medical Sciences, award R01GM121950; National Institute of Dental and Craniofacial Research, award number R01DE024468; and National Institute on Aging, award number UH2AG064704), the U.S. Department of Energy, Office of Science, Office of Biological & Environmental Research through the Microbial Community Analysis and Functional Evaluation in Soils SFA Program (m-CAFEs) under contract number DE-AC02-05CH11231 to Lawrence Berkeley National Laboratory, the National Science Foundation (grant NSFOCE-BSF 1635070), the Human Frontiers Science Program (RGP0020/2016), and the Boston University Interdisciplinary Biomedical Research Office.

>>>>

\section*{Acknowledgments}

We are grateful to members of the Segrè lab for helpful discussions and for feedback on the manuscript. This work was partially funded by grants from the National Institutes of Health (National Institute of General Medical Sciences, award R01GM121950; National Institute of Dental and Craniofacial Research, award number R01DE024468; and National Institute on Aging, award number UH2AG064704), the U.S. Department of Energy, Office of Science, Office of Biological \& Environmental Research through the Microbial Community Analysis and Functional Evaluation in Soils SFA Program (m-CAFEs) under contract number DE-AC02-05CH11231 to Lawrence Berkeley National Laboratory, the National Science Foundation (grant NSFOCE-BSF 1635070), the Human Frontiers Science Program (RGP0020/2016), and the Boston University Interdisciplinary Biomedical Research Office.

### 2020-08-25 sections/coverpage.tex
Updates from Overleaf

  \vspace{1.5 cm}
  Dileep Kishore\textsuperscript{1},
  Gabriel Birzu\textsuperscript{2},
  Zhenjun Hu\textsuperscript{1},
  Charles DeLisi\textsuperscript{1,3},
  Kirill Korolev\textsuperscript{$\dagger$1,2},\\
  Daniel Segr\`{e}\textsuperscript{$\dagger$1,3,4}\\
  \vspace{1cm}
  \textsuperscript{1}Bioinformatics Program, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{2}Department of Physics, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{3}Department of Biomedical Engineering, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{4}Department of Biology, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{$\dagger$}Correspondence should be sent to \href{mailto:korolev@bu.edu}{korolev@bu.edu} or \href{mailto:dsegre@bu.edu}{dsegre@bu.edu}\\

>>>>

  \vspace{1.5 cm}
  Dileep Kishore\textsuperscript{1,2},
  Gabriel Birzu\textsuperscript{3},
  Zhenjun Hu\textsuperscript{1},
  Charles DeLisi\textsuperscript{1,3},
  Kirill Korolev\textsuperscript{$\dagger$1,2,3},\\
  Daniel Segr\`{e}\textsuperscript{$\dagger$1,2,4,5}\\
  \vspace{1cm}
  \textsuperscript{1}Bioinformatics Program, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{2}Biological Design Center, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{3}Department of Physics, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{4}Department of Biomedical Engineering, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{5}Department of Biology, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{$\dagger$}Correspondence should be sent to \href{mailto:korolev@bu.edu}{korolev@bu.edu} or \href{mailto:dsegre@bu.edu}{dsegre@bu.edu}\\

### 2020-08-25 sections/introduction.tex
Updates from Overleaf

  Microbial communities are ubiquitous and play an important role in marine and terrestrial environments, urban ecosystems, metabolic engineering, and human health [REFs].
  These microbial communities, or microbiomes, often comprise several hundreds of different microbial strains interacting with each other and their environment, often through intricate metabolic and signaling relationships.
  Understanding how these interconnections shape community structure and functionalities is a fundamental challenge in microbial ecology, with applications in the study of microbial ecosystems across different biomes. 
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods,  more information can be extracted from these microbial community samples than ever before.
  In particular, high-throughput sequencing, including community metagenomic sequencing and sequencing of 16S rRNA gene amplicons, has the potential to help detect, identify and quantify a large portion of the constitutive microorganisms of a microbiome \cite{Jovel2016} [MORE REFs, Reviews, keystone papers...].
  These advances have led to large-scale data collection efforts involving environmental (\acl{emp}) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012}.

 This wealth of information on the composition and functions of a community at different times and under different environmental conditions has the potential to help us understand how communities assemble and operate.
 A powerful tool for translating microbiome data into knowledge is the construction of possible inter-dependence networks across species.
 The importance of these networks of relationships is two fold: first, such networks can serve as maps that help identify hubs of keystone species [REF], or basic microbiome changes that occur as a consequence of environmental perturbations or underlying host conditions [REF]; second, networks of inter-dependencies can serve as a key bridge towards building mechanistic models of microbial communities, greatly enhancing our capacity to understand and control them.
 For example, multiple studies have shown the importance of specific microbial interactions in the healthy microbiome \cite{Lloyd-Price2016} and others have shown how changes in these interactions can lead to dysbiosis \cite{Wang2017,Gilbert2016,Belizario2015}.
 In the context of terrestrial biogeochemistry..... [ANOTHER EXAMPLE].

 Direct measurement of interactions, e.g. through co-culture microdroplet experiments [REF Blainey], or spatial visualization of natural communities [REF Borisy]  is possible, but still challenging.
 In parallel, sequencing data across multiple samples can be used for estimating co-occurrence relationships between taxa.
 While the the relationship between directly measured interactions and statistically inferred co-occurrence is still poorly understood \cite{Zuniga2017}, a significant amount of effort has gone into estimating correlations from large microbiome sequence datasets.
 Co-occurrence networks have microbial taxa as nodes, and edges that represent the frequent co-occurrence (or negative correlations) across different datasets.

One of the most frequently used avenues for inferring co-occurrence networks is the parsing and analysis of 16S sequencing data. A large number of software tools and pipelines have been developed to analyze 16S sequencing data, often focused on addressing the many known limitations of this methodology, including resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.  Popular methods for different phases of the analysis of 16S data include; (i) tools for removing statistical artifacts...; (ii),

  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
  Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurrence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

 Previous work has developed popular platforms (like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita}) and tools (such as QIIME \cite{Caporaso2010}) that provide pipelines for 16S data analysis.

 However, existing tools typically are focused on \ac{otu} generation and not on the effects of upstream statistical methods on the inferred co-occurrence networks.
 Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data-analysis from amplicons to networks.

>>>>

  Microbial communities are ubiquitous and play an important role in marine and terrestrial environments, urban ecosystems, metabolic engineering, and human health \cite{Ghoul2016,Thompson2017}.
  These microbial communities, or microbiomes, often comprise several hundreds of different microbial strains interacting with each other and their environment, often through intricate metabolic and signaling relationships.
  Understanding how these interconnections shape community structure and functionalities is a fundamental challenge in microbial ecology, with applications in the study of microbial ecosystems across different biomes. 
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods,  more information can be extracted from these microbial community samples than ever before.
  In particular, high-throughput sequencing, including community metagenomic sequencing and sequencing of 16S rRNA gene amplicons, has the potential to help detect, identify and quantify a large portion of the constitutive microorganisms of a microbiome \cite{Jovel2016,Lloyd-Price2016}.
  These advances have led to large-scale data collection efforts involving environmental (\acl{emp}) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012}.

 This wealth of information on the composition and functions of a community at different times and under different environmental conditions has the potential to help us understand how communities assemble and operate.
 A powerful tool for translating microbiome data into knowledge is the construction of possible inter-dependence networks across species.
 The importance of these networks of relationships is two fold: first, such networks can serve as maps that help identify hubs of keystone species \cite{Menon2018,Rottjers2018}, or basic microbiome changes that occur as a consequence of environmental perturbations or underlying host conditions \cite{Gilbert2016}; second, networks of inter-dependencies can serve as a key bridge towards building mechanistic models of microbial communities, greatly enhancing our capacity to understand and control them.
 For example, multiple studies have shown the importance of specific microbial interactions in the healthy microbiome \cite{Lloyd-Price2016} and others have shown how changes in these interactions can lead to dysbiosis \cite{Wang2017,Gilbert2016,Belizario2015}.
 In the context of terrestrial biogeochemistry, co-occurrence networks have proposed as a valuable approach towards reconstructing the processes leading to microbiome assembly [REF Fierer], and understanding the response of microbial communities to environmental perturbations [REF].

 Direct high-throughput measurement of interactions, e.g. through co-culture microdroplet experiments [REF Blainey], or spatial visualization of natural communities [REF Borisy] is possible, but it requires specific technological capabilities, and has not been used very extensively yet.
 In parallel, sequencing data across multiple samples can be used for estimating co-occurrence relationships between taxa.
 While the the relationship between directly measured interactions and statistically inferred co-occurrence is still poorly understood \cite{Zuniga2017}, a significant amount of effort has gone into estimating correlations from large microbiome sequence datasets.
 Co-occurrence networks have microbial taxa as nodes, and edges that represent the frequent co-occurrence (or negative correlations) across different datasets.

One of the most frequently used avenues for inferring co-occurrence networks is the parsing and analysis of 16S sequencing data [REF]. A large number of software tools and pipelines have been developed to analyze 16S sequencing data, often focused on addressing the many known limitations of this methodology, including resolution, sequencing depth, compositional nature, sequencing errors and copy number variations [REF]. Popular methods for different phases of the analysis of 16S data include tools for: (i) denoising and clustering sequencing reads \cite{Caporaso2010,Callahan2016}; (ii) assigning taxonomy to the reads \cite{DeSantis2006,Quast2012}; (iii) processing and transforming the taxonomy count matrices \cite{Weiss2015}; and (iv) inferring the co-occurrence network \cite{Cougoul2019,Kurtz2015}.
Different specific algorithms are often aggregated into popular platforms (like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita}) and packages (such as QIIME \cite{Caporaso2010}) that provide pipelines for 16S data analysis.
The different methods and tools developed to solve issues arising in 16S analysis can lead to vastly different inferences of community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
 This is partially due to the fact that existing platforms are typically focused on \ac{otu} generation and not on the effects of upstream statistical methods on the inferred co-occurrence networks.
 Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data-analysis from amplicons to networks.
  More broadly, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurrence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.
